---
title: "How to vectorize a function"
author: "[Sanjay Bhatikar](https://in.linkedin.com/in/sanjaybhatikar)"
date: "`r Sys.Date()`"
output: html_document
---

# Problem Statement
You have packaged code in a function. Now you want to change the default behavior of the function to work with lists. This is consistent with the spirit of R. How do you magically endow your function to chow through lists?

# Solution Sketch
Use `Vectorize`. 

## Data Preparation
Priya, who manages my greenhouses in Bangalore, wants a weather forecast for the next 15 days. We grow a special variety of corn in the greenhouse. Her experience is that the number of cloudy days over the crop cycle affects yield. The more the number of cloudy days, the lower the yield. As we grow corn all year round, she can plan the greenhouse activities more effectively with a weather forecast sent daily to her over mail.

The Dark Sky API provides the weather at any time, anywhere in the world. To look up  the weather forecast for Bangalore, for example, see [here](http://forecast.io/#/f/12.9833,77.5833). I have written a function `getWeather` to fetch a forecast with a geo-coded location and time-stamp.

```{r results='hide'}
library(jsonlite)
source('keys') # Put your weather_key in this file.
```
```{r }
# Get weather object at a location and at a time
getWeather <- function(lat, long, timeStamp) {
  baseurl <- "https://api.forecast.io/forecast";
  weather_key # <- "get_your_own_dude"; 
  
  lat.long.time <- paste(lat, long, timeStamp, sep=",")
  sourceURL <- paste(baseurl, weather_key, lat.long.time, sep="/")
  
  sourceURL
  w <- fromJSON(sourceURL)
  w$daily$data$temperature <- w$currently$temperature
  w
}

```
Note that you need to register your app with the API, get a key and paste the string in the commented code for the function to run. Once past the preliminaries, execute the function as follows. 
```{r}
w <- getWeather(lat="12.9833", long="77.5833", timeStamp="2015-06-29T12:00:00+0530")
```
It generates the following weather forecast for Bangalore. Refer the API docs [here](https://developer.forecast.io/docs). 
```{r}
w$currently$summary
```
The forecast is *Mostly Cloudy* for that day. What if I have a list of geo-coded locations(lat-long coordinates with time-stamps) that I want to fetch weather data for? 

Locations |  Latitude | Longitude | Timestamp
----------|----------|-----------|----------
Abhayapuri |  26.40000 |	90.63333 |	2015-06-29T12:00:00+0530
Abohar | 30.16667 |	74.16667	| 2015-06-29T12:00:00+0530
Abu |	24.68333	| 72.83333	| 2015-06-29T12:00:00+0530
Achalpur	| 21.36667	| 77.53333	| 2015-06-29T12:00:00+0530
Acher	| 23.16667	| 72.53333	| 2015-06-29T12:00:00+0530
Adhoi	| 23.43333	| 70.53333	| 2015-06-29T12:00:00+0530
Adilabad	| 19.55000	| 78.33333	| 2015-06-29T12:00:00+0530
Adoni	| 15.55000	| 77.30000	| 2015-06-29T12:00:00+0530
Adra	| 23.50000	| 86.70000	| 2015-06-29T12:00:00+0530
Agar	| 23.06667	| 76.03333	| 2015-06-29T12:00:00+0530
Agartala	| 23.08333	| 91.38333	| 2015-06-29T12:00:00+0530

The obvious solution is to loop over the rows in the table. That can be clumsy and error-prone. The elegant solution is to vectorize the function.

## Execution
Use the `Vectorize` function.
```{r}
vecGetWeather <- Vectorize(getWeather, SIMPLIFY=FALSE)
```
The function is now ready to accept lists as inputs. That's all there is to it! In general, the output is a list of the same size as the input. 
```{r}
out <- vecGetWeather(lat=c(26.40000, 30.16667, 24.68333), 
                      long=c(90.63333, 74.16667, 72.83333), 
                      timeStamp="2015-06-29T12:00:00+0530")
```
Build on this illustration. Read the table into R. Supply the inputs as vectors.

## Output
1. **Input**: We have supplied three pairs of lat-long coordinates in separate lists per the interface. The lone time-stamp is the odd man out. It is recycled thrice. 
2. **Output**: It is a list of three items corresponding to the input. Each item is the result of a single API query. 

Note that:
1. The vectorized version of a function accepts lists and returns a list.
2. Where input lists are of unequal size, elements are recycled.
3. R expects vectorization as the default behavior from functions.

# Conclusion
Vectorization in R is as painless as it is awesome. It avoids the need for clumsily looping over rows of data. As a matter of fact, R expects vectorization as the default behavior of your functions. Use the simplicity and elegance of `Vectorize` to take your coding into pro territory. 
