---
title: "How to vectorize a function"
author: "[Sanjay Bhatikar](https://in.linkedin.com/in/sanjaybhatikar)"
date: "`r Sys.Date()`"
output: html_document
---

# Problem Statement
You have packaged code in a function. Now you want to change the default behavior of the function to work with lists. This is consistent with the spirit of R. How do you magically endow your function to chow through lists?

# Solution Sketch
Use `Vectorize`. 

## Data Preparation
Priya, who manages my greenhouses in Bangalore, wants a weather forecast for the next 15 days. We grow a special variety of corn in the greenhouse. Her experience is that the number of cloudy days over the crop cycle affects yield. The more the number of cloudy days, the lower the yield. She can plan the greenhouse activities more effectively with a weather forecast sent daily to her over mail.

The Dark Sky API provides the weather at any time, anywhere in the world. To look up  the weather forecast for Bangalore, for example, see [here](http://forecast.io/#/f/12.9833,77.5833). I have written a function to fetch a forcast.

```{r results='hide'}
library(jsonlite)
```
```{r }
# Get weather object at a location and at a time
getWeather <- function(lat, long, timeStamp) {
  baseurl <- "https://api.forecast.io/forecast";
  weather_key <- "get_your_own_dude"; 
  
  lat.long.time <- paste(lat, long, timeStamp, sep=",")
  sourceURL <- paste(baseurl, weather_key, lat.long.time, sep="/")
  
  sourceURL
  w <- fromJSON(sourceURL)
  w$daily$data$temperature <- w$currently$temperature
  w
}

```
Note that you need to register your app with the API, get a key and paste the string in the code for the function to run. Execute the funtion as follows. 
```{r}
w <- getWeather(lat="12.9833", long="77.5833", timeStamp="2015-06-29T12:00:00+0530")
```
It generates the following weather forecast for Bangalore. Refer the API docs [here](https://developer.forecast.io/docs). 
```{r}
w$currently$summary
```
What if I have a set of lat-long coordinates with time-stamps in a table that I want to fetch weather data for? I can loop over the rows in the table. A more elegant solution is to vectorize.

## Execution
Use the `Vectorize` function.
```{r}
vecGetWeather <- Vectorize(getWeather, SIMPLIFY=FALSE)
```

Locations |	Latitude | Longitude | Timestamp
----------|----------|-----------|----------
Abhayapuri |	26.40000 |	90.63333 |	2015-06-29T12:00:00+0530
Abohar | 30.16667 |	74.16667	| 2015-06-29T12:00:00+0530
Abu |	24.68333	| 72.83333	| 2015-06-29T12:00:00+0530
Achalpur	| 21.36667	| 77.53333	| 2015-06-29T12:00:00+0530
Acher	| 23.16667	| 72.53333	| 2015-06-29T12:00:00+0530
Adhoi	| 23.43333	| 70.53333	| 2015-06-29T12:00:00+0530
Adilabad	| 19.55000	| 78.33333	| 2015-06-29T12:00:00+0530
Adoni	| 15.55000	| 77.30000	| 2015-06-29T12:00:00+0530
Adra	| 23.50000	| 86.70000	| 2015-06-29T12:00:00+0530
Agar	| 23.06667	| 76.03333	| 2015-06-29T12:00:00+0530
Agartala	| 23.08333	| 91.38333	| 2015-06-29T12:00:00+0530

Can you think of putting the vectorized behavior of `vecGetWeather` to use?

## Output
Build on the illustration. Read the table into R. Supply the inputs as vectors.
```{r}
out <- vecGetWeather(lat=c(26.40000, 30.16667, 24.68333), 
                      long=c(90.63333, 74.16667, 72.83333), 
                      timeStamp="2015-06-29T12:00:00+0530")
```
**Input**: We have supplied three pairs of lat-long coordinates in separate lists per the interface. The lone time-stamp is the odd man out. It is recycled thrice. 
**Output**: It is a list of three items corresponding to the input. Each item is the result of a single API query. 

Note that:
* The vectorized version of a function accepts lists and returns a list.
* Where input lists are of unequal size, elements are recycled.
* R expects vectorization as the default behavior from functions.

# Conclusion
Vectorization in R is awesome. It avoids the need for looping over rows of data. R expects vectorization as the default behavior. Use the simplicity and elegance of `Vectorize` to take your functions into pro category. 
