---
title: "How To: Intersect Two Sets"
author: "[Sanjay Bhatikar](https://in.linkedin.com/in/sanjaybhatikar)"
date: "`r Sys.Date()`"
output: html_document
---

# Problem Statement
Find common items in two lists. This is equivalent to a Venn diagram.

# Solution Sketch
Use the operator `%in%` to find what two lists have in common.

## Data prepration
I want to find which actors from my list of favorites are also Oscar winners.
```{r}
favoriteActors <- c("Sean Connery", "Clint Eastwood", "Marlon Brando")
oscarWinners <- c("Marlon Brando", "Daniel Day-Lewis", "Tom Hanks", "Sean Penn", "Jack Nicholson")
```
## Execution
Which favorite actors are Oscar winners? Compare using `intersect`.
```{r echo=TRUE}
intersect(favoriteActors, oscarWinners)
```
What's left out? Compare using `setdiff`.
```{r echo=TRUE}
setdiff(favoriteActors, oscarWinners)
```
Compare using the `%in%` operator.
```{r echo=TRUE}
favoriteActors %in% oscarWinners
```
## Output
So only Marlon Brando makes the cut. Also see:
```{r echo=TRUE}
"Marlon Brando" %in% oscarWinners
```
No surprises there!

# Conclusion
Set operations for lists are quick and painless. Don't bother with loops. Instead, use any of the functions for set operations. 